The hidden manifold distance for functional data

### 2019-07-28 index.Rmd
some writing of sim scenario section

# Abstract
Functional data analysis is the statistical analysis of smooth infinite-dimensional curves. 
The challenge of analyzing infinite-dimensional objects is ameliorated by the fact that, since observations are smooth over their domain, the dimensionality of the curves is only artifically high.
Less studied in functional data analysis is the idea that the curves may actually live in a submanifold with low intrinsic dimension.
Interesting classes of functional manifold data include the class of probability density functions and the class of warped curves of a common template function. 
Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration. 
In this work we set ourselves the task of calculating the pairwise geodesic distance between functional observations that, due to noise, live near a non-linear manifold rather than exactly on the manifold. 
<!-- This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold.  -->
We will show that good estimation of the geodesic distance cannot be accomplished by naive methods such as smoothing followed by a shortest-path algorithm. 
Instead, we propose a dedicated technique for pairwise geodesic distance estimation by first projecting the functional data to the underlying hidden manifold and then perofrming operations on this hidden manifold. 
We conclude the work by showing the implications for many downstream tasks in functional data analysis, in particular distanced-based functional classification.

@@

Let $X_1,\ldots,X_n$ be a sample of $n$ independant realizations of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. We additionally assume that the function $X$ belongs to a low-dimensional nonlinear manifold $\M \subset L^2([a,b],\R)$, in which case a geodesic distance taking into account the intrinsic structure of $\M$ rather than the $L^2$ distance should be used as a measure of distance between two functions in $\M$. As many statistical methods depend on some measure of distance, downstream benefits of using the geodesic distance can be expected.

???insert example that shows pairwise geodesic distances give visually more interesting results than pairwise $L_2$ distance.???

@@

Given perfect functional observations, i.e.\ with no measurement error and on a very dense domain grid, the estimation of the geodesic distance can be accomplished straightforwardly using algorithms such as Floyd's. 
However, functional data are observed with noise in such a way that the observed functional data no longer live on a manifold. Though powerful techniques exist for smoothing discretely observed noisy functional data, the smoothing is designed for the $L_2$ recovery of the original curve, not to ensure the recovered functional versions of the data lie on or close to the functional manifold $\M$. 
<!-- This presents challenges to the application of many manifold learning techniques which assume fully observed nearly-noiseless inputs. -->

@@

## Simulation scenarios [Susan]
Acknowledge that we are sampling on a very concentrated measure to make sure functional manifold is well sampled. Cite Diaconis et. al's "Sampling from a manifold". How to sample properly from a functional manifold could be interesting future work.

* warping manifold (scenario 5 in sim_functional_data.R)
* pdf manifold (scenario 2 in sim_functional_data.R)
* square root pdf manifold (scenario 4 )

### Results

@@

For simplicity, assume the task is bianry classification. Associated to each functional object $x$ is a binary $y$ indicating class membership.
Consider the classifier proposed in Ferraty and Vieu (2003,2006) which is a functional version of the Nadaraya-Watson kernel estimator of class membership probabilities:
$$
\hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
$$
We shall compare our method to using $L_2$ distance, possibly weighted, and with curve registration already accomplished. Describe alternative methods in detail.

>>>>

# Abstract
Functional data analysis is the statistical analysis of smooth infinite-dimensional curves. 
The challenge of analyzing infinite-dimensional objects is ameliorated by the fact that, since observations are smooth over their domain, the dimensionality of the curves is only artifically high.
Less studied in functional data analysis is the idea that the curves may actually live in a submanifold with low intrinsic dimension.
Interesting classes of functional manifold data include the class of probability density functions and the class of warped curves of a common template function. 
Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration. 
In this work we set ourselves the task of calculating the pairwise geodesic distance between functional observations that, due to noise, live near a non-linear manifold rather than exactly on the manifold. Good estimation of the pairwise geodesic distance has beneficial implications for many downstream tasks in functional data analysis.
<!-- This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold.  -->
<!-- We will show that estimation of the geodesic distance cannot easily be accomplished by a naive method such as smoothing followed by a shortest-path algorithm. -->
The proposed methodology first projects the functional data on to the underlying hidden manifold and then performs operations on this hidden manifold. 
We conclude the work by using our pairwise geodesic distance estimator in distanced-based functional classification.

@@

Let $X_1,\ldots,X_n$ be a sample of $n$ independant realizations of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. We additionally assume that the function $X$ belongs to a low-dimensional nonlinear manifold $\M \subset L^2([a,b],\R)$. In this case a geodesic distance taking into account the intrinsic structure of $\M$ should be used as a measure of distance between two functions in $\M$, rather than the $L^2$ distance. As many statistical methods depend on some measure of distance, downstream benefits of using the geodesic distance can be expected.

<!-- ???insert example that shows pairwise geodesic distances give visually more interesting results than pairwise $L_2$ distance.??? -->

@@

Given perfect functional observations, i.e.\ with no measurement error and on a very dense domain grid, the estimation of the geodesic distance can be accomplished straightforwardly using algorithms such as Floyd's. 
However, when functional data are observed with noise, this approach is likely to fail. This is because we assume the error is such that the noisy functional manifold data no longer live on a manifold. 
Though powerful techniques exist for smoothing discretely observed noisy functional data, the smoothing is designed for the $L_2$ recovery of the original curve, not to ensure the recovered functional versions of the data lie on or close to the functional manifold $\M$. 
<!-- This presents challenges to the application of many manifold learning techniques which assume fully observed nearly-noiseless inputs. -->

@@

## Simulation scenarios [Susan]

### Isometric functional manifold of normal density functions
<!-- Scenario 5 -->
This scenario is modified from what is referred to as Manifold 2 in [@ChenMuller2012] by fixing the variance of the normal density to be $1$. We have $$\M =  \left \{X_\beta: \beta \in [-1,1], t \in [a,b]\right \}$$ with the $L_2$ inner product as the metric tensor of $\M$, where $X_\beta: [a,b] \to \R$ is given by $X_\beta(t) = \frac{1}{\sqrt{2\pi}} \exp{[-\frac{1}{2}(t-\beta)^2]}$. We set $a=-4$ and $b=4$.The geodesic distance between the curves $X_{\beta_1}$ and $X_{\beta_2}$ is given by
\begin{eqnarray*}
d(X_{\beta_1},X_{\beta_2}) &=& \int_{\beta_1}^{\beta_2} \left \| \frac{d X_\beta (t)}{d\beta} \right\|_{L^2} d\beta \\
&=&  \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4 \frac{1}{\sqrt{\pi}} \exp\{-(t-\beta)^2\}(t-\beta)^2 dt  }  \  d\beta \\
&=&   \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4(t-\beta)^2 f(t) dt  }  \  d\beta, \textrm{ where $f$ is the density of a N$(\beta,1/2)$ }   \\
&\approx&  \int_{\beta_1}^{\beta_2}  \sqrt{ \frac{1}{2\sqrt{\pi}} \frac{1}{2}} \ d\beta \\
&=& (\beta_2-\beta_1) \frac{1}{2\pi^{1/4}},
\end{eqnarray*}
where the approximation comes from the fact that we are integrating on $[a,b]=[-4,4]$ and not on $\R$. 
We can see this manifold is isometric, since the geodesic distance between $X_{\beta_1}$ and $X_{\beta_2}$ in $\M$ is the Euclidan distance between the $\beta$'s, up to some scaling factor. Note that the "straight" line connecting $X_{\beta_1}$ and $X_{\beta_2}$ in $\M$ does not always stay inside of $\M$, so we cannot employ the calculation technique of Scenario 1.

### Functional manifold of square root velocity functions
It was shown in [@Joshi2007] that the square root representation of probability density functions has a nice closed form geodesic.
They consider the manifold $$ \M = \{ \psi:[0,1] \to \R : \psi \ge 0, \int_0^1 \psi^2(s) \,ds = 1 \}$$ with the metric tensor given by the Fisher-Rao metric tensor $$ <v_1,v_2> = \int_0^1 v_1(s) v_2(s) \,ds $$ for two tangent vectors $v_1,v_2 \in T_\psi(\M)$. Note that this concides with the $L_2[0,1]$ inner product. [@Joshi2007] showed that the geodesic distance between any two $\psi_1$ and $\psi_2$ in $\M$ is simply $$d(\psi_1,\psi_2) = \cos^{-1}<\psi_1,\psi_2>.$$
We will specifically examine the square root of $Beta(\alpha,\beta)$ distributions which is supported on $[0,1]$. That is, $$ M = \{ \psi_{\alpha,\beta}: 1 \le \alpha \le 5, 2 \le \beta \le 5\} $$ where $\psi_{\alpha,\beta}: [0,1] \to \R$ is the pdf of $Beta(\alpha,\beta)$.

### Functional manifold of warping functions
<!-- Scenario 5 -->
This is based on Equations (17) and (18) of [@Kneip2008] but with $z_{i1}, z_{i2}$ set to $1$. (Equation 17 has a typo where the exponentials are missing negative signs).Let $X_\alpha(t) = \mu(h_\alpha(t))$ be defined on $[-3,3]$ where $$ \mu(t) = \exp\{(t-1.5)^2/2\} + \exp\{(t+1.5)^2/2\}$$ and $$ h_\alpha(t) = 6 \frac{ \exp\{\alpha(t+3)/6\} - 1}{\exp\{\alpha\}-1}, \alpha \ne 0 $$ and $h_\alpha(t) = t$ if $\alpha = 0$. Consider the manifold $$ M = \{X_\alpha: -1 \le \alpha \le 1\} $$. The geodesic distance is then $$ d(X_{\alpha_1},X_{\alpha_2}) = \int_{\alpha_1}^{\alpha_2} \left \| \frac{d X_\alpha (t)}{d\alpha} \right\|_{L^2} d\alpha.$$

### Sampling from a functional manifold
As of yet, there is little work as to how to sample from a functional manifold. Even in the Euclidean case, it is not obvious how sampling should be done [@Diaconis2013]. To safeguard against sampling unevenly on the functional manifold, we sample on a very concentrated measure for the intrinsic parameters.
For both the manifold of normal densities and the manifold of warping functions, we sample $\alpha$ according to ???rtruncnorm(samplesize,alphamin,alphamax,(alphamin+alphamax)/2,1)???
For the manifold of square root beta densities, we sample as follows       
alpha<- rtruncnorm(nb_alpha,alphamin,alphamax,(alphamax+alphamin)/2,0.3)
beta<- rtruncnorm(nb_beta,betamin,betamax,(betamax+betamin)/2,0.3)

How to sample properly from a functional manifold could be interesting future work.
### Results

@@

For simplicity, assume the task is binary classification. Associated to each functional object $x$ is a binary $y$ indicating class membership.
Consider the classifier proposed in Ferraty and Vieu (2003,2006) which is a functional version of the Nadaraya-Watson kernel estimator of class membership probabilities:
$$
\hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
$$
We shall compare our method to using $L_2$ distance, possibly weighted, and with curve registration already accomplished. Describe alternative methods in detail.
